A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-28 paper.tex
Fixed consistency of game trees in unbounded section

\tikzset{every node/.style={solid}}
\tikzstyle{fixed}=[solid]
\tikzstyle{unfixed}=[dash pattern = on 2pt off 2pt]
\begin{figure}
    \centering
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT}
        \label{fig:agt}
    \end{subfigure}%
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        node [left=4pt] {$\epsilon$}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    node [left=4pt] {$\beta$}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    node [left=4pt] {$\gamma$}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [right=4pt] {$\alpha$}
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT with partial strategy}
        \label{fig:strategy}
    \end{subfigure}%
    \begin{subfigure}[t]{.25\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right] {$e$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{Refined AGT}
        \label{fig:refined}
    \end{subfigure}
    \caption{Abstract game trees}
    \label{fig:alltrees}
\end{figure}

@@

The unbounded synthesis algorithm (Algorithm \ref{alg:unbounded}) effectively
consists of two communicating solvers: the counterexample-guided bounded
reachability solver and the unbounded solver based on incremental induction.
The unbounded solver calls the bounded solver with an increasing bound (line 8)
while learning states from abstract game trees that either must or may lose for
the controller. When the initial state is contained within the must-losing
states the game is known to be unrealisable (line 5).

May-losing states are learned by removing controller-winning states. We
maintain a set of states for every rank up the current bound. The states we
learn while executing the bounded algorithm can only said to be
controller-winning for a particular rank or below. However, we construct each
set carefully such that it is monotonically larger than the set of the rank
below (see Figure \ref{fig:maylosing}) and so that the environment is unable to
force play from one set to the next. Due to this careful construction, when two
adjacent sets become equivalent we know that the algorithm has reached a fixed
point and the controller is winning in the unbounded game (line 6).

\begin{figure}
    \centering
    \begin{tikzpicture}
        \node [minimum size=2cm,circle,draw] (a){$B^m[n]$};
        \node [minimum size=1.5cm,circle,draw,right=1.8cm of a] (b){$B^m[n-1]$};
        \node [minimum size=1cm,right=1.5cm of b] (dots){$\dots$};
        \node [minimum size=0.5cm,circle,draw,right=1.5cm of dots] (c){$B^m[0]$};
        \draw[-latex] (a.east) -- node [above] {$\supset$} (b.west);
        \draw[-latex] (b.east) -- node [above] {$\supset$} (dots);
        \draw[-latex] (dots.east) -- node [above] {$\supset$} (c.west);
    \end{tikzpicture}
    \caption{May-losing states}
    \label{fig:maylosing}
\end{figure}

@@

@@

\begin{figure}
    \centering
    \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
        \node [] (root){}
            child {node [circle,draw,label=left:$x_1$] {}
                child {node [circle,draw] {}
                    child {node [circle,draw,label=left:$x_0^0$] {} edge from parent[unfixed]}
                    edge from parent [fixed] node [left] {$u_1$}
                }
                child {node [circle,draw] {} edge from parent[unfixed]
                    child {node [circle,draw,label=right:$x_0^1$] {} edge from parent[unfixed]}
                    edge from parent [fixed] node [right] {$u_2$}
                }
                edge from parent [unfixed]
            };
    \end{tikzpicture}
%%%    \begin{tikzpicture}[overlay]
%%%        \draw [rotate = -20] (-1.9,0.35) ellipse (0.7 and 1.3);
%%%    \end{tikzpicture}
    \caption{A controller-losing game tree}
    \label{fig:interpolatetree}
\end{figure}

Consider the snippet of a game tree in Figure \ref{fig:interpolatetree}. The
tree is losing for the controller, the node labelled $x_1$ is at rank 1, and
$x_0^0$ and $x_0^1$ are at rank 0. Since the tree is controller-losing we know
that at least one run represented by the tree contains the error state.  As
a result, $\textsc{treeFormula}(gt)$ is unsatisfiable. If we take the step
$x_1$ to $x_0^0$ and cut it from the rest of the tree then
$\textsc{treeFormula}(step) \land \textsc{treeFormula}(parent)$ must too be
unsatisfiable.

We can construct an interpolant with $A = \textsc{treeFormula}(parent)$ and $B
= \textsc{treeFormula}(step)$. The only variables shared between $A$ and $B$
are the state variables at $x_1$. We know that $B \land \mathcal{I}$ is
unsatisfiable, therefore all states in $\mathcal{I}$ must lose to the
uncontrollable label $u_1$. We also know that $A \to \mathcal{I}$, thus
$\mathcal{I}$ contains all states reachable by the parent tree (on runs that
avoid the error state.)

@@

Now we can consider the step $x_1$ to $x_0^1$, and the parent - now without
either $x_1 \to x_0^0$ or $x_1 \to x_0^1$. The formula
$(\textsc{treeFormula}(parent) \land \mathcal{I}) \land
\textsc{treeFormula}(step)$ must be unsatisfiable. $\mathcal{I}$ contains all
states that lose to $u_1$ so any other state reachable at $x_1$ must lose to
$u_2$. Therefore we can compute another interpolant that contains states that
lose to $u_2$.

@@

We define two global invariants of the algorithm.  The \emph{may-invariant}
states that sets $B^m[i]$ grows monotonically with $i$ and that each $B^m[i+1]$
overapproximates the uncontrollable predecessor of $f^m[i]$: $$\forall
i<k.~B^m[i] \subseteq B^m[i+1], Upre(B^m[i]) \subseteq B^m[i+1].$$

>>>>

\tikzset{every node/.style={solid}}
\tikzstyle{fixed}=[solid]
\tikzstyle{unfixed}=[dash pattern = on 2pt off 2pt]
\begin{figure}
    \centering
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT}
        \label{fig:agt}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        node [left=4pt] {$\epsilon$}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    node [left=4pt] {$\beta$}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    node [left=4pt] {$\gamma$}
                    edge from parent [fixed] node [right] {$d$}
                }
                node [right=4pt] {$\alpha$}
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{AGT with partial strategy}
        \label{fig:strategy}
    \end{subfigure}%
    \begin{subfigure}[t]{.32\textwidth}
        \centering
        \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
            \node [circle,draw] (root){}
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [left] {$b$}
                    }
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {}
                    child {node [circle,draw] {}
                        edge from parent [fixed] node [right] {$e$}
                    }
                    edge from parent [fixed] node [right] {$d$}
                }
                node [above=4pt] {$\langle s, k \rangle$};
        \end{tikzpicture}
        \captionsetup{width=.8\textwidth}
        \caption{Refined AGT}
        \label{fig:refined}
    \end{subfigure}
    \caption{Abstract game trees}
    \label{fig:alltrees}
\end{figure}

@@

The unbounded synthesis algorithm enhances the bounded method with
computational learning. States are carefully recorded from unsatisfiable
abstract games while maintaining an invariant ensuring the monotonicity of
learning.

During the CEGAR loop of the bounded algorithm, each of the competing solvers
searches for satisfying assignments of labels to abstract game trees. When the
abstract game is unsatisfiable this indicates that the states represented by
nodes in the abstract game tree are losing for the current player. We extract
these states from the game tree using interpolation.

@@

The states in an abstract game with no controller candidate are
\emph{must-losing}. The controller will never be able to avoid the error state
from these states.

From abstractions with no environment candidate we record the complement states
as \emph{may-losing}. We maintain a set of states for every rank up the current
bound. The states we learn while executing the bounded algorithm can only said
to be controller-winning for a particular rank or below. However, we construct
each set carefully such that it is monotonically larger than the set of the
rank below
%(see Figure \ref{fig:maylosing})
and so that the environment is unable to force play from one set to the next.
Due to this careful construction, when two adjacent sets become equivalent we
know that the algorithm has reached a fixed point and the controller is winning
in the unbounded game (line 6).

%%%\begin{figure}
%%%    \centering
%%%    \begin{tikzpicture}
%%%        \node [minimum size=2cm,circle,draw] (a){$B^m[n]$};
%%%        \node [minimum size=1.5cm,circle,draw,right=1.8cm of a] (b){$B^m[n-1]$};
%%%        \node [minimum size=1cm,right=1.5cm of b] (dots){$\dots$};
%%%        \node [minimum size=0.5cm,circle,draw,right=1.5cm of dots] (c){$B^m[0]$};
%%%        \draw[-latex] (a.east) -- node [above] {$\supset$} (b.west);
%%%        \draw[-latex] (b.east) -- node [above] {$\supset$} (dots);
%%%        \draw[-latex] (dots.east) -- node [above] {$\supset$} (c.west);
%%%    \end{tikzpicture}
%%%    \caption{May-losing states}
%%%    \label{fig:maylosing}
%%%\end{figure}

@@

\begin{figure}
    \centering
%%%    \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
%%%        \node [circle,draw] (root){}
%%%            child {node [circle,draw] {}
%%%                child {node [circle,draw] {}
%%%                    edge from parent [fixed] node [left] {$b$}
%%%                }
%%%                edge from parent [fixed] node [left] {$a$}
%%%            }
%%%            child {node [circle,draw] {}
%%%                edge from parent [fixed] node [right] {$d$}
%%%            }
%%%            node [above=4pt] {$\langle s, k \rangle$};
%%%    \end{tikzpicture}
    \begin{tikzpicture}[dash pattern = on 2pt off 2pt, level distance = 10mm]
        \node [] (root){}
        child {node [circle,draw] {}
                child {node [circle,draw] {}
                    edge from parent [fixed] node [left] {$a$}
                }
                child {node [circle,draw] {} edge from parent[unfixed]
                    edge from parent [fixed] node [right] {$b$}
                }
            };
    \end{tikzpicture}
%%%    \begin{tikzpicture}[overlay]
%%%        \draw [rotate = -20] (-1.9,0.35) ellipse (0.7 and 1.3);
%%%    \end{tikzpicture}
    \caption{A controller-losing game tree}
    \label{fig:interpolatetree}
\end{figure}

Consider the snippet of a game tree in Figure \ref{fig:interpolatetree}. The
tree is losing for the controller, the topmost node is at rank 1, and the $a$ and
$b$ nodes are at rank 0. Since the tree is controller-losing we know that at
least one state in the tree will uncontrollably lead to the error state.  As a
result, $\textsc{treeFormula}(T)$ is unsatisfiable. If we take a step of the
tree, such as the step where the environment plays $a$, and cut it from the
rest of the tree then $\textsc{treeFormula}(step) \land
\textsc{treeFormula}(parent)$ must too be unsatisfiable.

We can construct an interpolant with $F_1 = \textsc{treeFormula}(parent)$ and
$F_2 = \textsc{treeFormula}(step)$. The only variables shared between $F_1$ and
$F_2$ are the state variables at rank 1. We know that $F_2 \land \mathcal{I}$
is unsatisfiable, therefore all states in $\mathcal{I}$ must lose to the
uncontrollable action $a$. We also know that $F_1 \to \mathcal{I}$, thus
$\mathcal{I}$ contains all states reachable by the parent tree (on runs that
avoid the error state.)

@@

Now we can consider the step where the environment plays $b$, and the parent
now without either step in the snippet. The formula
$(\textsc{treeFormula}(parent) \land \mathcal{I}) \land
\textsc{treeFormula}(step)$ must be unsatisfiable.  $\mathcal{I}$ contains all
states that lose to $a$ so any other state reachable at rank 1 in this subtree
must lose to $b$.  Therefore we can compute another interpolant that contains
states that lose to $b$.

@@

We define two global invariants of the algorithm.  The \emph{may-invariant}
states that sets $B^m[i]$ grows monotonically with $i$ and that each $B^m[i+1]$
overapproximates the uncontrollable predecessor of $B^m[i]$: $$\forall
i<k.~B^m[i] \subseteq B^m[i+1], Upre(B^m[i]) \subseteq B^m[i+1].$$
